---
title: "Results"
---

### Investigation and Selected Plots
We initially began by investigating the amount of marijuana use for 12-17 year olds in Colorado during the years surrounding its legalization in 2012 and comparing it to other substance use around the same times. Notably, marijuana use itself has been rapidly increasing on average over the past ten or so years.

```{r Colorado Time Series, echo = FALSE, message = FALSE}
library(tidyverse)
library(haven)
library(ggplot2)
library(plotly)

saes = read_sas("./state_saes_final.sas7bdat") %>%
  janitor::clean_names() %>% 
  filter(pyear > 4 & pyear != 12 & area == 2) %>% 
  select(outname, pyearnm, area, stname, pyear, agegrp, est_total, nsel, ncomp, pop) %>% 
  mutate(agegrp = ifelse(agegrp == 0, "All",
                         ifelse(agegrp == 1, "12 to 17",
                         ifelse(agegrp == 2, "18 to 25",
                         ifelse(agegrp == 4, "26 or older", NA))))) %>% 
  filter(!is.na(agegrp),
         outname %in% c("alcohol use in the past month", 
                        "binge alcohol use in the past month",
                      "illicit drug use other than marijuana in the past month",
                      "cigarette use in the past month",
                      "tobacco use in the past month",
                      "marijuana use in the past month"
                      )) 
saes$pyearnm = as.factor(saes$pyearnm)
levels(saes$pyearnm)[c(7)] <- c('2010-2011')


#Time-series graph: MJ use in Colorado among 12-17 yr olds
saes %>% 
    filter(outname == "marijuana use in the past month",
           agegrp == "12 to 17",
           stname == "Colorado") %>% 
    plot_ly(x = ~pyearnm, y = ~est_total, type = "scatter", mode = "lines+markers", hoverinfo = 'text', text = ~paste("Year:", pyearnm, '<br>', "Total count:", est_total),
          marker = list(size = 10,
                       color = 'rgba(255, 182, 193, .9)',
                       line = list(color = 'rgba(152, 0, 0, .8)',
                                   width = 2))) %>%
  layout(title = "Marijuana use in the past month among 12 - 17 year olds, Colorado",
         yaxis = list(zeroline = TRUE, title = "Estimated Total Number of Users (in thousands)"),
         xaxis = list(zeroline = TRUE, title = "Year Range"),
         margin = list(b = 100), xaxis = list(tickprefix = " "))
```

<br>
Unfortunately, it appeared that alcohol and illicit substance use was relatively unaffected in the years immediately following legalization. Legalization did not seem to affect the amount of people involving themselves with alcohol or illicit substances. Only cigarette use, visualized in the graph below, seemed to follow a downward trend that mirrored Marijuana's upward trend. While it would be fantastic to be able to suggest the two are related, most other states share a similar trend of decreasing cigarette use.

``` {r plots, echo = FALSE, message = FALSE}
#Time-Series graph: tobacco use in Colorado among 12-17 yr olds
saes %>% 
    filter(outname == "tobacco use in the past month",
           agegrp == "12 to 17",
           stname == "Colorado") %>% 
    plot_ly(x = ~pyearnm, y = ~est_total, type = "scatter", mode = "lines+markers", hoverinfo = 'text', text = ~paste("Year:", pyearnm, '<br>', "Total count:", est_total),
          marker = list(size = 10,
                       color = 'rgba(255, 182, 193, .9)',
                       line = list(color = 'rgba(152, 0, 0, .8)',
                                   width = 2))) %>%
  layout(title = "Tobacco use in the past month among 12 - 17 year olds, Colorado",
         yaxis = list(zeroline = TRUE, title = "Estimated Total Number of Users (in thousands)"),
         xaxis = list(zeroline = TRUE, title = "Year Range"),
         margin = list(b = 100), xaxis = list(tickprefix = " "))
```

<br>
We then decided to compare trends of marijuana use in Colorado with other states to examine if it had a higher rate of adoption following legalization. In the example shown below, we examined marijuana use for 18-25 year olds in Colorado compared to those in Alabama. From the comparison, it does appear that marijuana use in Colorado is increasing at a much faster rate in Colorado. It is difficult to say if this is attributable to legalization, given that it was increasing similarly before legalization.

```{r Comparison,  echo = FALSE, message = FALSE}
# Comparison graph: Colorado vs. Alabama, MJ use in past month, 18-25 yr olds

# Fix for ggplotly's poor graph axis scaling
levels(saes$pyearnm) = gsub("-", "\n \n", levels(saes$pyearnm))

# Generating ggplot
  p = saes %>%
    filter(stname %in% c("Colorado", "Alabama"), 
           agegrp == "18 to 25", 
           outname == "marijuana use in the past month") %>%
    ggplot(aes(x = pyearnm, y = est_total*1000/pop*100, fill = stname, text = paste('State:', stname, '</br></br>Year:', substring(pyearnm,1,4), '</br>Proportion Users:', round(est_total*1000/pop*100,2), "%"))) +
    geom_bar(colour = "black", position = position_dodge(width = 0.6), stat = "identity") + labs(
      title = "Comparing Monthly Marijuana Users Between States, ages 18 to 25",
      x = "Years",
      y = "Proportion Users to Population (%)",
      fill = "States Compared:"
    ) + 
    theme_bw() +
    theme(axis.text.x = element_text(vjust = 0.5), axis.title.x = element_blank(), legend.title = element_blank()) + 
        scale_y_continuous(limits = c(0, 100) ,breaks = seq(0, 100, 5));
  
  # Generating plotly graph. Attempt to make hovertext informative. Plotly does not handle hovertext well when creating the graph from ggplot
  ggplotly(p, tooltip = c('text'))

# I want to get off ggplotly's wild ride
levels(saes$pyearnm) = gsub("\n \n", "-", levels(saes$pyearnm))
```

<br><br>
Finally, to generate a broad visual of marijuana users compared to the general population, we generated a scatter plot plotting the two for selected years and age groups. In our selected example, we look at marijuana use for 12-17 year olds in 2014-2015 for all 50 states. There are a couple interesting points to consider when looking at this plot:

* It does not appear that total population correlates perfectly with users, as Texas has a greater population than New York with less marijuana users
* Most of the states that would vote to legalize marijuana in the future (Massachusets, Oregon, California) are over what appears to be the trendline for states

```{r Scatterplot, echo = FALSE, message = FALSE}
#Scatter graph: tobacco use in Colorado among 12-17 yr olds  
saes %>%
  filter(agegrp == "All", 
         outname == "marijuana use in the past month", 
         pyearnm == "2014-2015") %>%
  plot_ly(
  x = ~pop, y = ~est_total*1000,
  # Hover text:
  hoverinfo = 'text',
  text = ~paste("State:", stname, '<br>Total Users:', est_total*1000, '<br>Population:', pop),
  color = ~est_total, size = ~est_total) %>%
  layout(title = "Scatterplot Comparing MJ Users Against Total Population, 2014-2015",
         yaxis = list(title = "Estimated User Total"),
         xaxis = list(title = "Population Total "),
         margin = list(b = 100), xaxis = list(tickprefix = " "))
```

### Conclusions

From the data we have, it is difficult to make any suggestions that marijuana legalization is lowering use of illicit drugs or alcohol. However, marijuana use largely seems to be increasing while cigarette use is decreasing - though cigarette use decline is also seen in states without marijuana legalization. States that legalize marijuana tend to have a disproportionate number of users compared to the average of other states around the same population. Since this largely holds true even before legalization, it might suggest that those states with larger amounts of users are more likely to legalize marijuana, which would make sense. 

The effects of marijuana legalization will become clearer as more data becomes available. Additionally, more granular information on illicit drugs would help to inform us more as to marijuana legalization's effect on them. With more years and detailed data, future research into marijuana legalization will be able to provide a clearer picture regarding its public health effects.